Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2017-03-26 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
90 % Discussion

The next set of comparison were made with the natural gas burner fire positioned in the corner against two painted gypsum board walls. From the perspective of the impact of the corner on the visual fire behavior, the RI=20 simulation captures the extension of the combustion zone.  The similarity of the images in Figure~\ref{FDS_Corner_Sequence} and those in Figure~\ref{GBNG34Sequence} is very good.

@@

	\caption[Image sequence of an FDS/Smokeview simulation of the natural gas burner positioned in a corner.]{Image sequence of an FDS/Smokeview simulation of the natural gas burner positioned in a corner. The time interval between images is 0.3 of second.}
	\label{FDS_Corner_Sequence}
\end{figure}
The next comparison was between the Smokeview corner fire pattern images and photographs of corner fire patterns.  The images and photographs are presented in Figure~\ref{FDS_FirePattern_Corner_comp}.  The isotherms in the Smokeview images range from 20~$^\circ$C to greater than 360~$^\circ$C.  The `char' area with the threshold value of 360~$^\circ$C is shown in the Smokeview image on the right.  The upper edge of the char image was approximately 1.5~m above the burner.  As noted in Chapter 6, even though the combustion zone extended toward the ceiling, the flames appeared to move away from the walls as the flames moved away from the burner.  The natural gas generated corner fire patterns were similar in height to the natural gas burner generated wall patterns with an average height of 0.74~m.  The FDS value is approximately twice that.  Of course changing the ignition temperature to 520~$^\circ$C would bring the simulated char height close to the measured values.

@@

@@

@@

Compare flame height results IGWB with Fire Pattern Experiments   steep thermal gradient

@@

In these experiments, no significant differences were noted between the patterns generated between the open, closed and insulated construction methods for the walls.  The results from the instrumented gypsum wallboard experiments in Chapter 4, which represented an open construction type of wall, showed that the peak temperatures on the backside of the gypsum board were approximately 80~$^\circ$C.  The corresponding peak temperatures on the face of the gypsum wall board ranged from 415 to 540~$^\circ$C.  This demonstrates the insulation capability of the gypsum wallboard. Given the relatively low energy flow through the gypsum wallboard as represented by the limited temperature increase on the backside of the material.  Therefore with the limited fire exposure time of approximately 300~seconds and the small, pre-flashover fires used in this study, it follows that the additional insulation added with the closed and insulated construction methods would not have created temperature differences at the face of the gypsum wallboard.

Each of the two sets of experiments, the fire source characterization and the fire pattern, had uncertainties associated with them.  In the case of the fire source characterization, there were uncertainties associated with the fuel composition itself, the measurement of the heat release rate, temperature, heat flux and flame height.  The fire pattern experiments had additional uncertainties which included: the composition and thickness of the gypsum wallboard, moisture content and paint thickness.  In both sets of experiments, efforts were made to limit the uncertainties in the fuels, the measurement techniques and the gypsum wallboard. In this laboratory based study, the key question is – for a given fire pattern, is there a relationship that can be demonstrated to determine the source fire?

At this stage of the research, the answer to the above question is yes and no.  Figure 16 shows the comparisons of the average maximum fire pattern heights with those of the mean flame height based on the analysis of the video images.  The associated 95~\% confidence limits are also shown for each height.

@@

Screening algorithms utilizing the heat release rate results were employed to estimate the flame height and heat flux to a target surface.  These calculated values were compared to the measured results from “real fuels” and provided reasonable results.  Ranges of values were calculated in order to address the uncertainty in the heat release rate measurements.  Other uncertainties, such as lack of knowledge on the exact fuel composition need to be addressed, as this may well be the case in a real fire investigation.  Even accounting for uncertainties in heat release rate measurement, the flame height estimates tended to be better aligned with the peak flame height than the median flame height.  A specific coefficient for “gasoline” was used in Heskestad’s equation which resulted in the measured median flame height being within the range of the predictions.  The heat flux estimations were calculated with a range of radiant heat fractions, with the intent of the “actual” value being in that range.  With the exception of gasoline, the estimated heat flux values tended to under predict the measured results.  Given the limitations of the data, the algorithms provided reasonable estimates of the mean flame height and the heat flux.  These algorithms are tools which may be used to provide insight into a problem, but not an accurate single value answer.

\subsection{Modeling Results}

`Essentially, all models are wrong, but some are useful.'   Quote by George E.P. Box  statistician

@@

While you are reviewing the scene you meet up with the fire captain and begin asking him about the fire.  The captain tells you that the doors and windows were closed upon arrival.  When the front door was opened heavy smoke and flame 
flow path
inconsistent fire pattern near the right side of the sofa.  It appears that pattern was generated pre-flashover.  The pattern height is on the order of   calculate HRR and FH for cushion.
air equation

>>>>

The next set of comparisons were made with the natural gas burner fire positioned in the corner against two painted gypsum board walls. From the perspective of the impact of the corner on the visual fire behavior, the RI=20 simulation captures the extension of the combustion zone.  The similarity of the images in Figure~\ref{FDS_Corner_Sequence} and those in Figure~\ref{GBNG34Sequence} is very good.

@@

	\caption[Image sequence of an FDS/Smokeview simulation of the natural gas burner positioned in a corner.]{Image sequence of an FDS/Smokeview simulation of the natural gas burner positioned in a corner. The time interval between images is 0.3 of second.}
	\label{FDS_Corner_Sequence}
\end{figure}
The comparison between the Smokeview corner fire pattern images and photographs of corner fire patterns are presented in Figure~\ref{FDS_FirePattern_Corner_comp}.  The isotherms in the Smokeview images range from 20~$^\circ$C to greater than 360~$^\circ$C.  The `char' area with the threshold value of 360~$^\circ$C is shown in the Smokeview image on the right.  The upper edge of the char image was approximately 1.5~m above the burner.  As noted in Chapter 6, even though the combustion zone extended toward the ceiling, the flames appeared to move away from the walls as the flames moved away from the burner.  The natural gas generated corner fire patterns were similar in height to the natural gas burner generated wall patterns with an average height of 0.74~m.  The FDS value is approximately twice that.  Of course changing the ignition temperature to 520~$^\circ$C would bring the simulated char height close to the measured values.

@@

The fire patterns generated by the natural gas and gasoline fires had a high level of repeatability.  The 95~\% certainty range of the peak fire pattern heights, over 30 experiments from each fuel, was within approximately $\pm$0.1~m.  The maximum width range of the same experiments was within $\pm$0.07~m.  The ratio of the height to the width of the the pattern was 3 to 1.

The fire patterns formed by the polyurethane fires, which had an average peak heat release rate almost half of that of the natural gas and gasoline burner fires, had measured certainty ranges for height and width that were similar in magnitude ($\pm$0.12~m and $\pm$0.08~m) to the other two fire sources.  The polyurethane patterns where almost triangular in shape and had a height to width ratio of almost 1 to 1.  As a result the percentage difference between the average peak height of the pattern and range of of 95~\% certainty was approximately 3 times higher that the percentage differences from the natural gas and gasoline fires.

For all of the source fires the width of the patterns was similar to the width of the burner. The difference in the pattern height was due to the differences in heat release rate.  The average fire pattern areas of the natural gas and gasoline fires were approximately 0.15~m$^2$.  The average fire pattern area for the polyurethane foam fires was  0.05~m$^2$.  Given that the width of the fire pattern was tied to the burner width for this set of experiments, the ratio on the areas between the polyurethane foam and the other fuels was similar to the ratio of the heights of the fire patterns.

These experiments showed that the more repeatable the heat release rate, the more repeatable the fire pattern.

@@

The average peak height of the fire patterns from the natural gas burner, 0.73~m, was closely aligned with the median flame height of 0.71~m.  The average peak height of the fire patterns from gasoline pan fires was 0.82~m.  The median flame height for the gasoline was lower, 0.70~m. Examining the visible flame heights determined in Chapter 2, the average peak height of the fire pattern aligns better with the height where the gasoline flame was present for approximately 25~\% of the time.  The gasoline has a radiant fraction that is about twice as high as the natural gas.  This may account for the additional height on the gasoline fire patterns, relative to the natural gas fire patterns.

An observation from the experiments was that after a period of exposure time, the fire damage to the gypsum wall board would occur within seconds, and then the flame spread on the painted paper would stop and the pattern would not change size for the duration of the burn.  In one of the compartment fire pattern experiments the natural gas burner was allowed to burn for 15 minutes, yet the pattern did not increase after the initial formation.  This may be due to the steep thermal gradients that occurred at the line of demarcation between burned and not burned.

Both the fire source characterization and the compartment fire pattern experiments, had uncertainties associated with them.  In the case of the fire source characterization, there were uncertainties associated with the fuel composition itself, the measurement of the heat release rate, temperature, heat flux and flame height.  The fire pattern experiments had additional uncertainties or assumptions associated with the composition, thickness, and moisture content of the gypsum wall board. In both sets of experiments, efforts were made to limit the uncertainties in the fuels, the measurement techniques and the gypsum wallboard. one of the key question's was, is there a relationship that can be demonstrated to determine the source fire heat release rate based on the fire pattern?

Figure~\ref{FH_FP_comp} shows the comparisons of the average maximum fire pattern heights with those of the mean flame height based on the analysis of the video images.  The associated 95~\% confidence limits are also shown for each height.  For each of the fuels, the 95~\% certainty ranges of the fire pattern heights and flame heights overlap indicating that the values, while not an exact match, are similar.

\begin{figure}[h]
  \centering
  \includegraphics[width=5.0in]{../Figures/FH_FP_Comp}  \\
  \caption[Comparisons of the average maximum fire pattern heights compared with the mean flame heights.]{Comparisons of the average maximum fire pattern heights compared with the mean flame heights. The associated 95~\% confidence limits are also shown for each height.}
  \label{FH_FP_comp}
\end{figure}

@@

Comparing the limited number of fire pattern heights from the instrumented gypsum wall board experiments with the sets of 10 replicates from the compartment fire pattern experiments, show that the peak fire pattern heights were similar.  Based on these measurements, it would appear that the instrumentation did not impact the fire pattern results.

@@

In the compartment experiments, where 3 different wall construction methods were used, no significant differences were noted between the patterns generated between the open, closed and insulated construction methods for the walls.  The results from the instrumented gypsum wallboard experiments in Chapter 4, which represented an open construction type of wall, showed that the peak temperatures on the backside of the gypsum board were approximately 80~$^\circ$C.  The corresponding peak temperatures on the face of the gypsum wall board ranged from 415 to 540~$^\circ$C.  This demonstrated the insulative capability of the gypsum wall board. Given the relatively low energy flow through the gypsum wall board as represented by the limited temperature increase on the backside of the material.  Therefore with the limited fire exposure time of approximately 300~seconds and the small, pre-flashover fires used in this study, it follows that the additional insulation added with the closed and insulated construction methods would not have created significant temperature differences at the face of the gypsum wallboard.

\section{Modeling Results}

A few general thoughts to put the numerical methods in perspective.  In the experimental portion of this study, we established with the gas burner experiments that the steady state heat release rate values had a 95~\% certainty range of $\pm11\%$.  This range was the same as that associated with the heat release rate measurement system, so this would be the best agreement that we should expect.  The gasoline and the polyurethane introduced additional variability and the 95~\% certainty ranges on the peak heat release rate values increased to 30~\% and 47~\% respectively.  As the physical variability of the experiments increases what is the expectation of numerical methods which simulate the physical experiments

A statistician, George E.P. Box, is credited with the quote, `Essentially, all models are wrong, but some are useful.'  The ability of the numerical methods to simulate the results of the physical experiments was mixed.  Limited in some cases, by the available input data, inappropriate application, or lack of rigor in the underlying physics.  That said the models tended to provide enough inside to get a user `in the ballpark'.

\subsection{Algebraic Numerical Methods}

Screening algorithms utilizing the heat release rate results were employed to estimate the heat flux to a target surface and the flame height.  These calculated values were compared to the measured results from “real fuels” and provided reasonable results.  Ranges of values were calculated in order to address the uncertainty in the heat release rate measurements.  Other uncertainties, such as lack of knowledge on the exact fuel composition need to be addressed, as this may well be the case in a real fire investigation.  Even accounting for uncertainties in heat release rate measurement, the flame height estimates tended to be better aligned with the peak flame height than the median flame height.  A specific coefficient for “gasoline” was used in Heskestad’s equation which resulted in the measured median flame height being within the range of the predictions.  The heat flux estimations were calculated with a range of radiant heat fractions, with the intent of the “actual” value being in that range.  With the exception of gasoline, the estimated heat flux values tended to under predict the measured results.  Given the limitations of the input data, the algorithms provided reasonable estimates of the mean flame height and the heat flux.  These algorithms are tools which may be used to provide insight into a problem, but not an accurate single value answer.

@@

\subsection{Field Models}

The field model comparisons began with examining the vertical centerline temperatures and a remote heat flux measurement of the natural gas source fire with no walls present.  The simulation with a plume resolution index of 5 (RI=5) provided the best agreement. Best agreement in this case meaning that the predicted temperature and heat flux values were within the experimental uncertainty of the measured values. The flame height calculated using the HRRPUL provided similar results to the measured values.   Other RIs were tried and provided results that provided less agreement, therefore it seems critical that end users ensure that they have chosen a grid size that best represents the scale of fire flow field being modeled.

Once the wall was introduced, the flame and plume of the fire were displaced off of the vertical centerline toward the wall.  The amount of displacement in the physical experiments exceeded the simulated displacement.  This resulted in poor agreement, meaning simulated values outside of the uncertainty bounds of the measured values.  Once the burner was positioned against the wall the simulated values and the measured temperature values converged.

The flame and plume movement issue carried through the predicted conditions on the wall surface for both the temperature and the centerline heat flux values.  The simulated values trended with the measured values but the level of agreement between was mixed.  The worst agreement was observed at the vertical heat flux array positioned on the edge of the burner.  As the burner was moved closer to the wall the agreement between the measurement and the predicted heat flux be came worst.  This was due to the steep thermal gradient along the vertical edge as evidenced by the fire pattern line of demarcation between the burned and unburned paper on the gypsum wall board.  In the physical experiments the flames moved over the sensors, while in the simulation the flames seem to be offset from the sensors which resulted in under-predicted heat flux values especially at the positions closer to the wall.

The numerical methods are useful tools for gaining insight into fire behavior but in many cases many would require specific experimental data for input or for validation before the results could be relied upon for fire investigation use.

@@

In Chapter 1, as the investigator, you had completed your review of the scene and had collected evidence which included an electric heater.  You are not sure sure about the fire's progression or the area of origin.  Before leaving the scene, you meet up with the fire captain and begin asking him about the fire.  The captain tells you that the doors and windows of the townhouse were closed upon arrival.  When the front door was opened heavy smoke flowed out, followed by flames.  Other fire fighters were venting the upstairs windows.  The fire grew and burned for a short time before water was applied to the   
flow path
inconsistent fire pattern near the right side of the sofa.  It appears that pattern was generated pre-flashover.  The pattern height is on the order of   calculate HRR and FH for cushion.
air equation
